---
title: "Arizona School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Arizona School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  eval = TRUE
)
```

```{r load-packages}
library(azschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores Arizona's public school enrollment data.

> **Note:** Currently available for years 2018, 2024, and 2025. Years 2019-2023 are not available as Excel downloads.

---

## Statewide Enrollment Trends

```{r statewide-trend}
enr <- fetch_enr_multi(c(2018, 2024, 2025), use_cache = TRUE)

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#BF0A30") +
  geom_point(size = 3, color = "#BF0A30") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Arizona Public School Enrollment (2018-2024)",
    subtitle = "Growth has slowed in recent years",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

## Gender Distribution

```{r gender-by-year}
gender_by_year <- enr |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) |>
  select(end_year, subgroup, n_students) |>
  pivot_wider(names_from = subgroup, values_from = n_students)

gender_by_year
```

```{r gender-chart}
enr |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("male", "female")) |>
  mutate(pct = n_students / n_students[subgroup == "total_enrollment"][1] * 100,
         subgroup = ifelse(subgroup == "male", "Male", "Female")) |>
  ggplot(aes(x = end_year, y = pct, fill = subgroup)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Male" = "#0072B2", "Female" = "#CC79A7")) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    title = "Gender Distribution Over Time",
    x = "School Year",
    y = "Percentage",
    fill = "Gender"
  )
```

## Data Availability

This package uses data from the Arizona Department of Education October 1 enrollment reports.

**Currently available years:** 2018, 2024, 2025

**Missing years:** 2019-2023 (Excel files not available as automated downloads - Cloudflare protection blocks programmatic access)

**Data included:**
- State totals by grade level and gender
- District and school totals
- County-level aggregates

For more information, see the package README and function documentation.
